Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-05-20 pkg/glmertree/paper/treatment_interactions.Rnw
Minor improvements in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@858 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\linespread{1.2}
\usepackage{tabto}

@@

<<artificial-data>>=
set.seed(123)
treatment <- rbinom(n=150,size=1,prob=1/2)
duration <- round(matrix(rnorm(150,mean=5,sd=3),ncol=1))
duration[duration<0] <- 1
anxiety <- round(rnorm(150,mean=10,sd=3))
age <- round(rnorm(150,mean=30,sd=10))
age[age<18] <- 18
age[age>75] <- 75
cutscore1 <- 5+.44*3
cutscore2 <- 10
node1t1 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==0, 1, 0)
node1t2 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==1, 1, 0)
node2t1 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==0, 1, 0)
node2t2 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==1, 1, 0)
node3t1 <- ifelse(duration>cutscore1 & treatment==0, 1, 0)
node3t2 <- ifelse(duration>cutscore1 & treatment==1, 1, 0)
beta1 <- -2
beta2 <- 0
beta3 <- 2.5
error <- rnorm(150,0,2)
cluster <- error+rnorm(150, 0, 6)
int_values <- rnorm(10, 0, 1)
tmp <- rep(int_values,each=15)
tmp <- tmp[order(tmp)]
rand_int <- rep(NA, times=150)
rand_int[order(cluster)] <- tmp
cluster[order(cluster)] <- rep(1:10, each=15)
cor(error,rand_int)
true_error <- error - rand_int
true_eff <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
     + .4*treatment)
outcome <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
                      + .4*treatment + true_error + rand_int)
outcome[outcome<0] <- 1
example <- data.frame(treatment,outcome,duration,anxiety,cluster, true_eff)
example$treatment <- factor(example$treatment,labels=c("Treatment 1","Treatment 2"))
#cor(outcome,rand_int)^2
#cor(outcome, true_eff)^2
#truetree <- lmtree(true_eff ~ treatment | age + duration + anxiety, data=example)
@

@@

@@

  \item[Step 2:] Estimate random effects in the mixed effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
  % Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

For the estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). This provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.

>>>>

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{bigfoot}
\usepackage{ctable}
\usepackage{rotating}
\usepackage{lineno}
\usepackage{placeins}
\linespread{1.2}
\usepackage{tabto}

@@

<<artificial-data>>=
set.seed(123)
treatment <- rbinom(n=150,size=1,prob=1/2)
duration <- round(matrix(rnorm(150,mean=5,sd=3),ncol=1))
duration[duration<0] <- 1
anxiety <- round(rnorm(150,mean=10,sd=3))
age <- round(rnorm(150,mean=30,sd=10))
age[age<18] <- 18
age[age>75] <- 75
cutscore1 <- 5+.44*3
cutscore2 <- 10
node1t1 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==0, 1, 0)
node1t2 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==1, 1, 0)
node2t1 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==0, 1, 0)
node2t2 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==1, 1, 0)
node3t1 <- ifelse(duration>cutscore1 & treatment==0, 1, 0)
node3t2 <- ifelse(duration>cutscore1 & treatment==1, 1, 0)
beta1 <- -2
beta2 <- 0
beta3 <- 2.5
error <- rnorm(150,0,2)
cluster <- error+rnorm(150, 0, 6)
int_values <- rnorm(10, 0, 1)
tmp <- rep(int_values,each=15)
tmp <- tmp[order(tmp)]
rand_int <- rep(NA, times=150)
rand_int[order(cluster)] <- tmp
cluster[order(cluster)] <- rep(1:10, each=15)
cor(error,rand_int)
true_error <- error - rand_int
true_eff <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
     + .4*treatment)
outcome <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
                      + .4*treatment + true_error + rand_int)
outcome[outcome<0] <- 1
example <- data.frame(treatment, outcome, duration, anxiety, cluster, age)
example$treatment <- factor(example$treatment, labels = c("Treatment 1", "Treatment 2"))
@

@@

The \verb|R| code for running the illustrative examples is provided in the Appendix.

@@

  \item[Step 2:] Estimate random effects in the mixed-effects model $x_{i}^{\top}\hat{\beta}_{j(r)} + z_{i}^{\top}\hat{b}_{(r)}$ with subgroups $j(r)$ from the GLM tree.
  % Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$, with $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.

@@

R \cite{R14} was used for generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| (or \verb|glmer|, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

For the estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge. This provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.
